### 2021-08-04 response.tex

Thank you for the review. 
We have tried to make the limitation and assumptions clearer throughout the text.

\revtext{1. (page 3, first paragraph) It is important to acknowledge that the renewal equation approach to the relationship between the generation interval
distribution, R, and the growth rate r assumes a mass-action epidemic.
When we assume a fixed generation interval distribution, we must add the
assumption that the rate of depletion of susceptibles over time is negligible and the prevalence of infection is low. One of the most important
findings of Ref 10 is that the mean generation interval contracts when the
rate of depletion of susceptibles increases. While the renewal equation
does allow for the generation interval to change over time, the analyses
in Sections 3–5 assume fixed mean generation intervals. Because variants
are likely to spread when susceptibles are being depleted by both variants
(due to cross-protection) and when the prevalence of infection is not low, these implicit assumptions place significant limitations on the practical
implications of the analysis.}

We now acknowledge that we are assuming homogeneous mixing---we feel that this is a clearer explanation because mass-action is often used to refer to how the transmission rate, beta, scales with the population size.

@@

We note that we primarily focus on short-term dynamics until Section 4.
In this case, we can approximate the dynamics using exponential growth/decay, and therefore the generation-interval distribution is expected to remain fixed:

``Over a short period of time, we can assume that epidemiological conditions remain roughly constant: $\RR_x(t) \approx \RR_x$ and $g_x(t, \tau) \approx g_x(\tau)$, in which case the incidence of infections changes exponentially.
Here, we use the term ``epidemiological conditions'' to broadly refer to all factors that affect transmission---mathematically, they are captured by the kernel $K(t, \tau)$.
In the context of SARS-CoV-2 infections, we are essentially assuming that the changes in susceptible pool, behavior, and contact rates are usually small over a short period of time.
Then, the incidence of each strain grows (or decays) exponentially at rate $r_x$, satisfying the Euler-Lotka equation \citep{wallinga2007generation}:
\begin{equation}
\frac{1}{\RR_x} = \int_0^\infty \exp(- r_x \tau) g_x(\tau) \dtau.
\end{equation}''

We further note that our fixed generation-interval distribution does not rely on the assumption that the rate of depletion of susceptibles over time is negligible.
This is because we are assuming that intrinsic, not realized, generation intervals are fixed.
As shown in Champredon et al 2015 and 2018, the standard SEIR model can be written as a renewal equation using a fixed \emph{intrinsic} generation-interval distribution, and we can still observe the contraction of the \emph{realized} generation interval for this model.
We have added a paragraph in Section 2 to clarify this distinction (please see response to reviewer 1 for text).

### 2021-08-05 newvariant.tex
downing

Inferring the relative strength (i.e., the ratio of reproduction numbers) and relative speed (i.e., the difference between growth rates) of new SARS-CoV-2 variants compared to the wild types, from which they emerged, is critical to predicting and controlling the course of the current pandemic.
We explore how differences in the generation-interval distributions affect the inference of relative strength and speed in light of recent evidence that the Alpha and Delta variants may have different generation intervals from prior dominant lineages.
Neglecting these differences gives biased estimates of relative strength, which can vary systematically with changes in the incidence of infections caused by the variant.
We further argue that holding relative strength constant is generally more appropriate for estimating the transmission advantage of new SARS-CoV-2 variants by comparing how relative strength and speed change over time under two idealized interventions of constant strength and speed.
Constant-strength interventions like idealized vaccination and social distancing, which reduce transmission rates by a constant amount, affect relative speed but not relative strength.
Constant-speed interventions like idealized contact tracing, which isolate infected individuals at a constant rate, affect relative strength but not relative speed.
Changes in the generation-interval distribution can alter the effectiveness of these interventions.

@@

Epidemic speed can be typically estimated from time series of incidence of infection during the exponential growth period \citep{mills2004transmissibility,nishiura2009transmission,ma2014estimating}, but epidemic strength is difficult to measure from incidence time series.
Instead, epidemic strength is often inferred from the observed epidemic speed using the generation-interval distribution, an approach popularized by \citep{wallinga2007generation}.
The generation interval, defined as the time between infection and transmission, provides information about the time scale of individual-level transmission \citep{svensson2007note}.
The generation interval is also distinctly different from other so-called ``transmission intervals'' that measure time between successive infections---including the serial interval, which is defined as the time between symptom onsets in an infector-infectee pair \citep{fine2003interval,grassly2008mathematical,britton2019estimation,ali2020serial,park2021forward}.

The exact shape of the distribution depends on several factors---including the shape of latent and infectious period distributions \citep{lloyd2001realistic,wearing2005appropriate,roberts2007model} as well as more detailed life history of a disease \citep{huber2016quantitative}---and thus can be difficult to estimate.
While it is possible to consider general forms of generation-interval distributions \citep{miller2010epidemics,svensson2015influence}, summarizing the distribution in terms of its mean and variability---for example, by assuming they are Gamma distributed---can still provide a robust link between epidemic speed and strength for real diseases and yield important biological insights \citep{park2019practical}.
In particular, several studies have noted, in many contexts, that mechanisms that increase the mean generation interval increase epidemic strength $\RR$ given observed epidemic speed $r$ \citep{eaton2014proportion,powers2014impact,weitz2015modeling,park2020time}.

Analyses of new variants have typically characterized \emph{relative} strength (i.e., the ratio of reproduction numbers $\Rv/\Rw$, here called $\rho$) and speed (i.e., the difference between growth rates $\rv-\rw$, here called $\delta$) of the variants.
Many studies have tried to estimate the relative strength of variants from the observed relative speed \citep{davies2021estimated, leung2021early, volz2021transmission,zhao2021,campbell2021increased}.
Instead (or additionally), some studies have assumed a value of the relative strength and tried to predict its relative speed \citep{davies2021estimated,di2021impact}.
While both approaches are reasonable, holding different quantities constant (i.e., strength or speed) can lead to different conclusions about the spread of the disease and its control because the two quantities are linked by the generation-interval distribution \citep{doi:10.1098/rspb.2020.1556}.

For example, when epidemic strength $\RR$ is fixed, assuming longer generation intervals results in a slower estimated epidemic growth rate (lower $r$), making an epidemic look easier to control with a constant-speed intervention, which isolates infected individuals at a constant hazard---such intervention affects late transmission disproportionately and further affect the post-intervention generation-interval distribution.
As shown in \cite{doi:10.1098/rspb.2020.1556}, the hazard of isolation must be greater than the epidemic speed $r$ to reach the $r=0$ threshold.
In this same case, however, variation in the generation-interval distribution has no effects on the effectiveness of a constant-strength intervention, which reduces the hazard of infectious contacts and therefore affects transmission by an equal amount regardless of age of infection.
Conversely, when epidemic speed $r$ is fixed, assuming longer generation intervals leads to a higher estimate of epidemic strength ($\RR$), making the epidemic look harder to control with a constant-strength intervention.
Constant strength and speed interventions are idealized representations of real-life interventions, which can be either strength-like (e.g., vaccination and social distancing) or speed-like (e.g., contact tracing and isolation) depending on how their effectiveness varies across the generation interval (see \cite{doi:10.1098/rspb.2020.1556} and Discussion).

@@

Unsurprisingly, we find that an increased speed of $\delta=0.1\pday$ for the variant is consistent with higher strength than the wild type across the range of epidemiological conditions considered (\fref{relstrength}).
However, the magnitude of relative strength $\rho$ is sensitive to assumptions about generation intervals:
For realistic values of $\kappa$ (excluding 0 and 1), the inferred relative strength $\rho$ ranges between 1.1 and 2.3 when $\Gv$ is allowed to vary between $2\Gw/3$ and $3\Gw/2$.

>>>>

Inferring the relative strength (i.e., the ratio of reproduction numbers) and relative speed (i.e., the difference between growth rates) of new SARS-CoV-2 variants compared to earlier strains is critical to predicting and controlling the course of the current pandemic.
We explore how differences in the generation-interval distributions affect the inference of relative strength and speed in light of recent evidence that the Alpha and Delta variants may have different generation intervals from prior dominant lineages.
Neglecting these differences gives biased estimates of relative strength, which can vary systematically with changes in the incidence of infections caused by the variant.
We further argue that holding relative strength constant is generally more appropriate for estimating the transmission advantage of new SARS-CoV-2 variants by comparing how relative strength and speed change over time under two idealized interventions of constant strength and speed.
Constant-strength interventions like idealized vaccination and social distancing, which reduce transmission rates by a constant amount, affect relative speed but not relative strength.
Constant-speed interventions like idealized contact tracing, which isolate infected individuals at a constant rate, affect relative strength but not relative speed.
Changes in the generation-interval distribution can alter the effectiveness of these interventions.

@@

Epidemic speed can be typically estimated from time series of incidence of infection during the exponential growth period \citep{mills2004transmissibility,nishiura2009transmission,ma2014estimating}, but epidemic strength is difficult to measure from incidence time series.
Instead, epidemic strength is often inferred from the observed epidemic speed using the generation-interval distribution, an approach popularized by \citep{wallinga2007generation}.
The generation interval, defined as the time between infection and transmission, provides information about the time scale of individual-level transmission \citep{svensson2007note}.
The generation interval is also distinctly different from other ``transmission intervals'' that measure time between successive infections---including the serial interval, which is defined as the time between symptom onsets in an infector-infectee pair \citep{fine2003interval,grassly2008mathematical,britton2019estimation,ali2020serial,park2021forward}.

The exact shape of the distribution depends on several factors---including the shape of latent and infectious period distributions \citep{lloyd2001realistic,wearing2005appropriate,roberts2007model} as well as more detailed life history of a disease \citep{huber2016quantitative}---and thus can be difficult to estimate.
While it is possible to consider general forms of generation-interval distributions \citep{miller2010epidemics,svensson2015influence}, summarizing the distribution in terms of its mean and variability---for example, by assuming they are Gamma distributed---can still provide a robust link between epidemic speed and strength for real diseases and yield important biological insights \citep{park2019practical}.
In particular, several studies have noted, in many contexts, that mechanisms that increase the mean generation interval increase the epidemic strength $\RR$ that corresponds to an observed epidemic speed $r$ \citep{eaton2014proportion,powers2014impact,weitz2015modeling,park2020time}.

Analyses of new variants have typically characterized \emph{relative} strength (i.e., the ratio of reproduction numbers $\Rv/\Rw$, here called $\rho$) and speed (i.e., the difference between growth rates $\rv-\rw$, here called $\delta$) of the variants.
Many studies have tried to estimate the relative strength of variants from the observed relative speed \citep{davies2021estimated, leung2021early, volz2021transmission,zhao2021,campbell2021increased}.
Instead (or additionally), some studies have assumed a value of the relative strength and tried to predict relative speed \citep{davies2021estimated,di2021impact}.
While both approaches are reasonable, holding different quantities constant (i.e., strength or speed) can lead to different conclusions about the spread of the disease and its control\citep{doi:10.1098/rspb.2020.1556}.

For example, when epidemic strength $\RR$ is fixed, assuming longer generation intervals results in a slower estimated epidemic growth rate (lower $r$), making an epidemic look easier to control with a constant-speed intervention, which isolates infected individuals at a constant hazard---such intervention affects late transmission disproportionately and further affect the post-intervention generation-interval distribution.
As shown in \cite{doi:10.1098/rspb.2020.1556}, the hazard of isolation must be greater than the epidemic speed $r$ to reach the $r=0$ threshold.
In this same case, however, variation in the generation-interval distribution has no effects on the effectiveness of a constant-strength intervention, which reduces the hazard of infectious contacts and therefore affects transmission by an equal amount regardless of age of infection.
Conversely, when epidemic speed $r$ is fixed, assuming longer generation intervals leads to a higher estimate of epidemic strength ($\RR$), making the epidemic look harder to control with a constant-strength intervention.
Constant-strength and -speed interventions are idealized representations of real-life interventions, which can be either strength-like (e.g., vaccination and social distancing) or speed-like (e.g., contact tracing and isolation) depending on how their effectiveness varies across the generation interval (see \cite{doi:10.1098/rspb.2020.1556} and Discussion).

@@

Unsurprisingly, we find that an increased speed of $\delta=0.1\pday$ for the variant is consistent with higher strength than the wild type across the range of epidemiological conditions considered (\fref{relstrength}).
However, the magnitude of relative strength $\rho$ is sensitive to assumptions about generation intervals:
For realistic values of $\kappa$ (excluding 0 and 1), the inferred relative strength $\rho$ ranges between 1.1 and 2.3 when $\Gv$ is allowed to vary between $2/3$ and $3/2$ of \Gw.
